Evolution of ORM Systems: Formal Foundations

### 2013-01-07 doc/description.tex
formating and minor fixes

\section{ORM Software}
\label{sec:evoIntro}
\begin{figure}
\centering
	\includegraphics[scale=0.3]{./images/system}
	\caption{The model of one generation of an ORM system consists of application persistent layer and relational database for storing instances.}
\label{fig:appStructure}
\end{figure}
A software implemented in an object-oriented language, which uses relational database as a data storage is context of this paper. The situation is illustrated in the Figure \ref{fig:appStructure}. There are three important components of such a software: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM (we will use $\rho$ as a symbol for ORM), therefore we define a software a triple:
\begin{gather*}
	software = < Application, Database, \rho > \\
	sw : Applications \times Databases \times ORM \rightarrow Software \\ \\
	application : Software \rightarrow Application \\
	application(sw(a, d, \rho)) = a \\ \\
	database : Software \rightarrow Database \\
	database(sw(a, d, \rho)) = d \\
	a \in Applications, d \in Databases
\end{gather*}
The software has to be in consistent state so it is beneficial for its user. The state of the software is defined to communicate the sate of the software to its user:
\begin{gather*}
	state = Consistent\; s \: | \: Inconsistent \; s, \\
	s \in Softwares  \\ \\
	state : Software \rightarrow State \\
	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases}
\end{gather*}
\subsection{Software Evolution}
\begin{figure}
\centering
	\includegraphics[scale=0.4]{./images/evolution_simple}
	\caption{The evolution of data changes the system on all levels. The figure shows all the components of the evolution process.}
	\label{fig:evolution}
\end{figure}
The evolution of a software is a transformation from one consistent state to another one. These states are called generations of the software. The evolution consists of change in application and in database as shown in the Figure \ref{fig:evolution}. We focus on situations when the evolution is initiated on the application level (as code refactoring), however the evolution has to be processed on the database level too. It means there is a set of application transformations called $AppOps$ and a set of database transformations called $DbOps$. Because there are needed different parameters for application and database evolution the set of transformation called $E$ which evolves the whole software and a mapping between $E$ and $AppOps$, respectively $DbOps$ is defined. We call this mappings $\Psi$ and $\Phi$:
\begin{gather*}
\Psi : E \rightarrow AppOps \\
\Phi : E \rightarrow DbOps
\end{gather*}
A subset of $E$ represents an evolution transformation of a software, which mets the first goal of our framework - a single source of data evolution for the whole software.

The application of a transformation on a software state (evolution) is marked by the $\oplus$ operator:
$$\oplus : State \times E \rightarrow State $$
and similarly the $\circ$ and the $\bullet$ operators for evolving application and database:
\begin{gather*}
\circ : Application \times AppOps \rightarrow Application \\
 \bullet : Database \times DbOps \rightarrow Database
\end{gather*}
The definition of the evolving operator is inspired by the monads (as used e.g. in Haskell programming language) so the state of the software can be passed from one evolution step to another, which helps us to concatenate evolution transformations.

\begin{gather*}
 u \oplus t = \begin{cases}
 	\inference{u = Inconsistent \; s}{u} \\ \\
 	\inference{u = Consistent \; s}{state(sw(\Psi(u, t), \Phi(u,t))} 
\end{cases}\\ 
u \in States, t \in E, s \in Softwares
\end{gather*}
\begin{gather*}
 a \circ o = \begin{cases}
 	\inference{!\;consistent(a)}{a} \\ \\
 	\inference{consistent(a)}{a \circ o} 
\end{cases}\\ 
a \in Applications, o \in AppOps
\end{gather*}
\begin{gather*}
 d \bullet o = \begin{cases}
 	\inference{!\;consistent(d)}{d} \\ \\
 	\inference{consistent(d)}{d \circ o} 
\end{cases}\\ 
d \in Applications, o \in DbOps
\end{gather*}

@@

\subsection{Application Model}
\label{sec:appModel}
The model represents a persistent layer of an application created using object oriented language. It means the layer is a set of instances which cooperates together. Each instance is defined according to a class - the class defines the structure of its instances and relationships between them an the rest of the application. The instances are stored in the relational database in our case, therefore we define only the structure of classes in our model. Besides the structure definition, there is defined a set of functions, which allows to manipulate the application model easily.

\begin{gather*}
  Application = < Class* > \\ 
  app : Class* \rightarrow  Application \\ \\
  classes : Application \rightarrow Class* \\
  classes(app(cs)) = cs  \\\\
  class : Label \times Application \rightarrow Class   \\ 
  class(l, a) = c, c \in classes(a) \wedge name(c) = l \\ \\
  l \in Label, a = Application, cs \in Class* 
  \\ \\
  associated :  Label \times Application \rightarrow Boolean \\
 \end{gather*}

\paragraph{Class} represents a basic structural concept in the application model. It has a unique name, one or more properties and a class can be associated to other classes in the application.

$$Class = <label, Property*, Association*>$$
\begin{equation*}
\begin{alignat}{2}
cl : Label \times Property* \times Association* \rightarrow Class &\quad&\text{constructor}\\ \\
% TODO association has to be only to self (all associated has to exists)
name : Class \rightarrow Label &\quad&\text{name of the class}\\
name(cl(l, props, assocs)) = l \\ \\
properties : Class \rightarrow Property* &\quad&\text{all properties}\\
properties(cl(l, props, assocs)) = props \\ \\
primitiveProperties : Class \rightarrow Property* &\quad&\text{all primitive properties}\\
primitiveProperties(cl(l, props, assocs)) = props', \\ props' \subset props,  \forall p \in props' : cardinality(p) \leq 1 \\ \\
collectionProperties : Class \rightarrow Property* &\quad&\text{all collections}\\
collectionProperties(cl(l, props, assocs)) = props', \\ props' \subset props, \forall p \in props' : cardinality(p) > 1 \\ \\
associations : Class \rightarrow Association* &\quad&\text{all associations}\\
associations(cl(l, props, assocs)) = assocs
\end{alignat}
\end{equation*}
$$l \in Label, props \in Properties, assocs \in Associations$$

\paragraph{Property} represents a feature of  a class which is represented as a primitive type. The property can be mandatory, can have a default value and can represent a single value or a collection of values.
\begin{gather*}
Property = <label, AppType, DefaultValue, Cardinality, Mandatory> \\ \\
prop : Label \times AppType \times String \times \mathbb{N^{*}} \times Boolean \rightarrow Property \\ \\
name : Property \rightarrow Label \\
name(prop(l, t, val, n, m)) = l \\ \\
type : Property \times AppType \\
type(prop(l, t, val, n, m)) = t \\ \\
cardinality : Property \rightarrow \mathbb{N^{*}} \\
cardinality(prop(l, t, val, n, m)) = n \\ \\
mandatory : Property \rightarrow Boolean \\
mandatory(prop(l, t, val, n, m)) = m  \\ \\
l \in Label, t \in AppType, val \in String*, n \in \mathbb{N^{*}}, m \in Boolean
\end{gather*}

@@

\paragraph {Association} represents a connection between two classes. It has a unique name and it contains a name of the class which is referenced by the association. The class which owns the association is consider to be the starting class of an association, referenced class is consider to be the ending class of an association. The cardinalities defines the multiplicity of the association ends.
\begin{gather*}
Association = <label, classRef, startCardinality, endCardinality> \\ \\
assoc : Label \times Label \times \mathbb{N} \times \mathbb{N} \rightarrow Association
name : Association \rightarrow Label \\
name(assoc(l, ref, n_1, n_2)) = l\\ \\
reference : Association \rightarrow Label \\
reference(assoc(l, ref, n_1, n_2)) = ref\\ \\
startCardinality : Association \rightarrow \mathbb{N} \\
startCardinality(assoc(l, ref, n_1, n_2)) = n_1\\ \\
endCardinality : Association \rightarrow \mathbb{N} \\
endCardinality(assoc(l, ref, n_1, n_2)) = n_2 \\ \\
l, ref \in Label,  n_1, n_2 \in \mathbb{N}
\end{gather*}

@@

\paragraph{AppType} represents primitive types in the application. There are usually defined types such as String, Integer, Boolean etc. in contrast there is only one type in our model, because we focus on structural and data changes and type casting operations are not important for us. The only App-Type type is called APP-STRING.
$$
AppType = APPSTRING
$$

\subsection{Database Model}
\label{sec:dbModel}
The relational database consists of database schema which defines the structure of the database and data, which in our ORM system represents instances. The database is:
\begin{gather*}
Database = <table*, row*> \\ \\
db : Tables \times Rows \rightarrow Database \\ \\
tables : Database \rightarrow Tables \\
tables(db(tabs, rows)) = tabs \\ \\
data : Database \rightarrow Rows \\
data(db(tabs, rows)) = rows \\ \\
table : Label \times Database \rightarrow Table  \\
table(l, d) = t, t \in tables(d) \wedge name(t) = l \\ \\
tabs \in Tables, rows \in Rows, l \in Labels, d \in Databases
\end{gather*}

\subsubsection{Database Schema Model}
The database schema model is defined by following concepts:
\paragraph{Table} represents a basic concept of database schema. It has a unique name, one or more column and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
\begin{gather*}
Table = <label, primaryKey, Column*, ForeignKey*>\\ \\
tab : Label \times PrimaryKeys \times Columns \times ForeignKeys \rightarrow Table \\ \\
name : Table \rightarrow Label  \\
name(tab(l, pk, cols, fks)) = l \\ \\
primaryKey : Table \rightarrow PrimaryKeys  \\
primaryKey(tab(l, pk, cols, fks)) = pk \\ \\
columns : Table \rightarrow Columns  \\
columns(tab(l, pk, cols, fks)) = cols \\ \\
foreignKeys : Table \rightarrow ForeignKeys  \\
foreignKeys(tab(l, pk, cols, fks)) = fks \\ \\
l \in Label, pk \in PrimaryKey, cols \in Columns, fks \in ForeignKeys
\end{gather*}

\paragraph{Column} columns define possible data values and types which can be part of a table record.
\begin{gather*}
Column = <label, DbType, DefaultValue, Constraint*> \\ \\
col : Label \times DbType \times String \times Constraints \rightarrow Column \\ \\
name : Column \rightarrow Label  \\
name(col(l, t, val, cons)) = l  \\ \\
type : Column \rightarrow DbType  \\
type(col(l, t, val, cons)) = t  \\ \\
constraints : Column \rightarrow Constraints  \\
constraints(col(l, t, val, cons)) = cons  \\ \\
l \in Labels, t \in DbType, val \in Strings, cons \in Constraints
\end{gather*}

\paragraph{Foreign key}
\begin{gather*}
ForeignKey = <label, tableRef, Constraint*> \\ \\
fk : Label \times Label \times Constraints \rightarrow ForeignKey \\ \\
name : ForeignKey \rightarrow Label \\
name(fk(l, tRef, cons)) = lab  \\ \\
reference : ForeignKey \rightarrow Label  \\
reference(fk(l, tRef, cons)) = tRef  \\ \\
reference : ForeignKey \rightarrow Constraint*  \\
reference(fk(l, tRef, cons)) = cons  \\ \\
l, tRef \in Label, cons \in Constraints
\end{gather*}

\paragraph{Primary key}
\begin{gather*}
PrimaryKey =  < label > 	\\
pk : Label \rightarrow PrimaryKey
\end{gather*}

\paragraph{DbType} represents primitive types in the database. There are usually defined types such as Varchar, Integer, Boolean etc. There are only two types defined in the model $DBSTRING$ as a default type of values in columns and $DBINT$ as a default type for keys.
$$
DbType = DBSTRING \; | \; DBINT 
$$

\paragraph{Constraint} there are two types of constraints defined in the model. Both constraints are column constraints - first constraint defines unique records in a column, second constraint defines non-empty columns.
\begin{gather*}
Constraint = NOTNULL \; | \; UNIQUE 
\end{gather*}

@@

\subsubsection{Model of Stored Data}
A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of reference to a concrete table and a set of value pairs, which represents concrete values of concrete columns.
\begin{gather*}
row = < refTable, pair* > \\
row : Label \times Pairs \rightarrow Row \\ \\
refTable : Row \rightarrow Label \\
refTable(row(l, pairs) = l \\ \\
contains : Table \times Database \rightarrow Boolean \\
contains(t, d) = \begin{cases}
 \inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
  \inference{otherwise}{false}
 \end{cases} \\ \\
 l \in Labels, pairs \in Pairs, t \in tables(d), d \in Database
\end{gather*}
There functions $contains$ verify if the table contains data in context of concrete database.

\begin{gather*}
pair = < refCol, value > \\
[Label \times Value] \rightarrow Pair\\ \\
\mathbf{pairByRef} : Label \times Row \rightarrow Pair \\
pairByRef(l,i) = \begin{cases}
	\inference{\exists p \in pairs(i) : reference(p) = l}{p} \\ \\
	\varnothing
 \end{cases}\\\\
\mathbf{reference} : Pair \rightarrow Label \\
reference([l,val]) = l \\ \\
\mathbf{value} : Pair \rightarrow Label \\ 
value([l,val]) = val \\
l \in Label, val \in Value
\end{gather*}
The $refCol$ is a name of a column, a foreign key or its primary key and $value$ represents concrete value stored in the database. Each row in a database has to refer to a table in database schema and has a structure corresponding to the one defined by a table, otherwise the database is inconsistent. The consistency of database is then:
\begin{gather*}
\mathbf{consistent} : Database \rightarrow Boolean \\
consistent(d) = \begin{cases}
 	\inference{\forall r \in data(d), \exists t \in tables(d) : reference(r) = name(t) \\ \wedge \forall p \in pairs(r) :\\ \exists u \in columns(t) \cup foreignKeys(t) : reference(p) = name(u)}{true}
 	\\ \\
 	\inference{otherwise}{false}
 \end{cases}
\end{gather*}

@@

\begin{gather*}
addColumn : Label \times Column* \times Database \rightarrow Database \\ addColum(l, c, d) = \begin{cases}
\inference{|c| > 1}{addColumn(l, head(c), d) \bullet addColumn(l, tail(c), d)}
\\ \\
\inference{|c| = 1 \wedge \exists t \in tables(d) \wedge name(t) = l \\ \wedge \forall f \in columns(t) : name(f) \neq name(c)}{columns(t) \cup c}
\\ \\
Inconsistent \;d 
 \end{cases}
\\ l \in Label, c = Column*, d = Database
\end{gather*}

\begin{gather*}
	copyColumn : Label \times Label \times Label \times Database \rightarrow Database \\
	copyColumn(l_s, l_t, l_c, d) = \begin{cases}
 		\inference{\exists t \in tables(d) : name(t) = l_s \\ \wedge \exists c \in columns(t) : name(c) = l_c \\ \wedge \exists e \in tables(d) : name(e) = l_t  \\ \wedge \forall q \in properties(e) : name(q) \neq l_c}{columns(e) \cup c }
	\end{cases} \\
	l_s,l_t,l_c \in Label, d = Database
\end{gather*}

\begin{gather*}
addTable : Table* \times Database \rightarrow Database \\ 
addTable(t, d) = \begin{cases}
	\inference{|t| > 1}{addTable(head(t),d) \bullet addTable(tail(t), d)}
\\ \\	
	\inference{|t| = 1 \wedge \forall e \in tables(d) \wedge name(t) \neq name(e)}{tables(d) \cup c}
\\ \\
Inconsistent \;d 
 \end{cases}
\\ t = Table*,  d = Database
\end{gather*}

\begin{gather*}
insertRow : Row* \times Database \rightarrow Database \\
insertRow(r, d) = \begin{cases}
	\inference{|r|>1}{insertRow(head(r), d) \bullet insertRow(tail(r), d)} 
\\ \\
 	\inference{|r|=1 \wedge \exists t \in tables(d) : name(t) = reference(r) }{data(d) \cup r}
\\ \\
 Inconsistent \; d
 \end{cases}
\\ r = Row*, d = Database
\end{gather*}

\begin{gather*}
appendValues : Label \times Pair* \times Database \rightarrow Daatabase \\
appendValues(l, p, d) = \begin{cases}
 	\inference{|p| > 1}{appendValues(l, head(p), d)\bullet appendValues(l, tail(p), d)} 
 \\ \\
 	\inference{|p| = 1 \wedge \exists r \in data(d), reference(r) = l}{pairs(r) \cup [name(p), defaultValue(p)]}
 \\ \\
 	 Inconsistent \; d
 \end{cases} 
 \\
l \in Label, p = Pair*, d = Databases
\end{gather*}

\begin{gather*}
insertPair : Row \times Pair \times Database \rightarrow Database \\
insertPair(r_t, p, d) = \begin{cases}
 \inference{\forall e \in pairs(r_t) : reference(e) \neq reference(p) \\ \wedge \exists t \in tables(d) : name(t) = refernces(r_t), \\ \wedge \exists c \in columns(t) : reference(p) = name(c)}{ pairs(r_t) \cup p}
 \\\\
 Inconsistent \; d
 \end{cases}
\end{gather*}

>>>>

\section{ORM Software}
\label{sec:evoIntro}
\begin{figure}
\centering
	\includegraphics[scale=0.3]{./images/system}
	\caption{The model of one generation of an ORM system consists of application persistent layer and relational database for storing instances.}
\label{fig:appStructure}
\end{figure}
A software implemented in an object-oriented language, which uses relational database as a data storage is context of this paper. The situation is illustrated in the Figure \ref{fig:appStructure}. There are three important components of such a software: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM (we will use $\rho$ as a symbol for ORM), therefore we define a software a triple:
\begin{align*}
	software = < Application, Database, \rho > \\
	sw : Applications \times Databases \times ORM \rightarrow Software \\ \\
	application : Software \rightarrow Application \\
	application(sw(a, d, \rho)) = a \\ \\
	database : Software \rightarrow Database \\
	database(sw(a, d, \rho)) = d \\
	a \in Applications, d \in Databases
\end{align*}
The software has to be in consistent state so it is beneficial for its user. The state of the software is defined to communicate the sate of the software to its user:
\begin{align*}
	state = Consistent\; s \: | \: Inconsistent \; s, \\
	s \in Softwares  \\ \\
	state : Software \rightarrow State \\
	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases}
\end{align*}
\subsection{Software Evolution}
\begin{figure}
\centering
	\includegraphics[scale=0.4]{./images/evolution_simple}
	\caption{The evolution of data changes the system on all levels. The figure shows all the components of the evolution process.}
	\label{fig:evolution}
\end{figure}
The evolution of a software is a transformation from one consistent state to another one. These states are called generations of the software. The evolution consists of change in application and in database as shown in the Figure \ref{fig:evolution}. We focus on situations when the evolution is initiated on the application level (as code refactoring), however the evolution has to be processed on the database level too. It means there is a set of application transformations called $AppOps$ and a set of database transformations called $DbOps$. Because there are needed different parameters for application and database evolution the set of transformation called $E$ which evolves the whole software and a mapping between $E$ and $AppOps$, respectively $DbOps$ is defined. We call this mappings $\Psi$ and $\Phi$:
\begin{align*}
\Psi : E \rightarrow AppOps \\
\Phi : E \rightarrow DbOps
\end{align*}
A subset of $E$ represents an evolution transformation of a software, which mets the first goal of our framework - a single source of data evolution for the whole software.

The application of a transformation on a software state (evolution) is marked by the $\oplus$ operator:
$$\oplus : State \times E \rightarrow State $$
and similarly the $\circ$ and the $\bullet$ operators for evolving application and database:
\begin{align*}
\circ : Application \times AppOps \rightarrow Application \\
 \bullet : Database \times DbOps \rightarrow Database
\end{align*}
The definition of the evolving operator is inspired by the monads (as used e.g. in Haskell programming language) so the state of the software can be passed from one evolution step to another, which helps us to concatenate evolution transformations.

\begin{align*}
 u \oplus t = \begin{cases}
 	\inference{u = Inconsistent \; s}{u} \\ \\
 	\inference{u = Consistent \; s}{state(sw(\Psi(u, t), \Phi(u,t))} 
\end{cases}\\ 
u \in States, t \in E, s \in Softwares
\end{align*}
\begin{align*}
 a \circ o = \begin{cases}
 	\inference{!\;consistent(a)}{a} \\ \\
 	\inference{consistent(a)}{a \circ o} 
\end{cases}\\ 
a \in Applications, o \in AppOps
\end{align*}
\begin{align*}
 d \bullet o = \begin{cases}
 	\inference{!\;consistent(d)}{d} \\ \\
 	\inference{consistent(d)}{d \circ o} 
\end{cases}\\ 
d \in Applications, o \in DbOps
\end{align*}

@@

\subsection{Application Model}
\label{sec:appModel}
The application model represents a persistent layer of an application created using object oriented language and transformations which can alter the model. 
\subsubsection{Application Structure}
Each instance is defined according to a class - the class defines the structure of its instances and relationships between them an the rest of the application. The instances are stored in the relational database in our case, therefore we define only the structure of classes in our model. Besides the structure definition, there is defined a set of functions, which allows to manipulate the application model easily.

\begin{align*}
  \mathbf{Application} = < Class* > \\ 
  \mathbf{app} : Class* \rightarrow  Application \\ \\
  \mathbf{classes} : Application \rightarrow Class* \\
  classes(app(cs)) = cs  \\\\
  \mathbf{class} : Label \times Application \rightarrow Class   \\ 
  class(l, a) = c, c \in classes(a) \wedge name(c) = l \\ \\
  l \in Label, a = Application, cs \in Class* 
  \\ \\
  \mathbf{associated} :  Label \times Application \rightarrow Boolean \\
\end{align*}

\paragraph{Class} represents a basic structural concept in the application model. It has a unique name, one or more properties and a class can be associated to other classes in the application. 
\begin{align*}
	\mathbf{Class} = <label, Property*, Association*> \\
	\mathbf{cl} : Label \times Property* \times Association* \rightarrow Class \\\\
% TODO association has to be only to self (all associated has to exists)
	\mathbf{name} : Class \rightarrow Label \\
	name(cl(l, props, assocs)) = l \\ \\
	\mathbf{properties} : Class \rightarrow Property* \\
	properties(cl(l, props, assocs)) = props \\ \\
	\mathbf{primitiveProperties} : Class \rightarrow Property* \\	primitiveProperties(cl(l, props, assocs)) = props', \\ props' \subset props,  \forall p \in props' : cardinality(p) \leq 1 \\ \\
	\mathbf{collectionProperties} : Class \rightarrow Property* \\	collectionProperties(cl(l, props, assocs)) = props', \\ props' \subset props, \forall p \in props' : cardinality(p) > 1 \\ \\
	\mathbf{associations} : Class \rightarrow Association* \\	associations(cl(l, props, assocs)) = assocs \\ \\
l \in Label, props \in Properties, assocs \in Associations
\end{align*}

\paragraph{Property} represents a feature of  a class which is represented as a primitive type. The property can be mandatory, can have a default value and can represent a single value or a collection of values.
\begin{align*}
	\mathbf{Property} = <label, AppType, DefaultValue, Cardinality, Mandatory> \\ \\
	\mathbf{prop} : Label \times AppType \times String \times \mathbb{N^{*}} \times Boolean \rightarrow Property \\ \\
	\mathbf{name} : Property \rightarrow Label \\
	name(prop(l, t, val, n, m)) = l \\ \\
	\mathbf{type} : Property \times AppType \\
	type(prop(l, t, val, n, m)) = t \\ \\
	\mathbf{cardinality} : Property \rightarrow \mathbb{N^{*}} \\
	cardinality(prop(l, t, val, n, m)) = n \\ \\
	\mathbf{mandatory} : Property \rightarrow Boolean \\
	mandatory(prop(l, t, val, n, m)) = m  \\ \\
l \in Label, t \in AppType, val \in String*, n \in \mathbb{N^{*}}, m \in Boolean
\end{align*}

@@

\paragraph {Association} represents a connection between two classes. It has a unique name and it contains a name of the class which is referenced by the association. The class which owns the association is consider to be the starting class of an association, referenced class is consider to be the ending class of an association. The cardinalities defines the multiplicity of the association ends.
\begin{align*}
	\mathbf{Association} = <label, classRef, startCardinality, endCardinality> \\ \\
	\mathbf{assoc} : Label \times Label \times \mathbb{N} \times \mathbb{N} \rightarrow Association\\
	\mathbf{name} : Association \rightarrow Label \\
	name(assoc(l, ref, n_1, n_2)) = l\\ \\
	\mathbf{reference} : Association \rightarrow Label \\
	reference(assoc(l, ref, n_1, n_2)) = ref\\ \\
	\mathbf{startCardinality} : Association \rightarrow \mathbb{N} \\
	startCardinality(assoc(l, ref, n_1, n_2)) = n_1\\ \\
	\mathbf{endCardinality} : Association \rightarrow \mathbb{N} \\
	endCardinality(assoc(l, ref, n_1, n_2)) = n_2 \\ \\
l, ref \in Label,  n_1, n_2 \in \mathbb{N}
\end{align*}

@@

\paragraph{AppType} represents primitive types in the application. There are usually defined types such as String, Integer, Boolean etc. in contrast there is only one type in our model, because we focus on structural and data changes and type casting operations are not important for us. The only App-Type type is called APP-STRING.
\begin{align*}
\mathbf{AppType} = APPSTRING
\end{align*}

\subsubsection{Application Transformation}

\subsection{Database Model}
\label{sec:dbModel}
The relational database consists of database schema which defines the structure of the database and data, which in our ORM system represents instances. The database is:
\begin{align*}
	\mathbf{Database} = <table*, row*> \\ \\
	\mathbf{db} : Tables \times Rows \rightarrow Database \\ \\
	\mathbf{tables} : Database \rightarrow Tables \\
	tables(db(tabs, rs)) = tabs \\ \\
	\mathbf{rows} : Database \rightarrow Rows \\
	rows(db(tabs, rs)) = rs \\ \\
	\mathbf{table} : Label \times Database \rightarrow Table  \\
	table(l, d) = t, t \in tables(d) \wedge name(t) = l \\ \\
	tabs = Table*, rs = Row*, l \in Label, d = Database
\end{align*}

\subsubsection{Database Schema Model}
The database schema model is defined by following concepts:
\paragraph{Table} represents a basic concept of database schema. It has a unique name, one or more column and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
\begin{align*}
	\mathbf{Table} = <label, primaryKey, Column*, ForeignKey*>\\ \\
	\mathbf{tab} 	: Label \times PrimaryKeys \times Columns \times ForeignKeys \rightarrow Table \\ \\
	\mathbf{name} : Table \rightarrow Label  \\
	name(tab(l, pk, cols, fks)) = l \\ \\
	\mathbf{primaryKey} : Table \rightarrow PrimaryKeys  \\
	primaryKey(tab(l, pk, cols, fks)) = pk \\ \\
	\mathbf{columns} : Table \rightarrow Columns  \\
	columns(tab(l, pk, cols, fks)) = cols \\ \\
	\mathbf{foreignKeys} : Table \rightarrow ForeignKeys  \\
	foreignKeys(tab(l, pk, cols, fks)) = fks \\ \\
l \in Label, pk = PrimaryKey, cols = Column*, fks = ForeignKey*
\end{align*}

\paragraph{Column} columns define possible data values and types which can be part of a table record.
\begin{align*}
	\mathbf{Column} = <label, DbType, DefaultValue, Constraint*> \\ \\
	\mathbf{col} : Label \times DbType \times String \times Constraints \rightarrow Column \\ \\
	\mathbf{name} : Column \rightarrow Label  \\
	name(col(l, t, val, cons)) = l  \\ \\
	\mathbf{type} : Column \rightarrow DbType  \\
	type(col(l, t, val, cons)) = t  \\ \\
	\mathbf{constraints} : Column \rightarrow Constraints  \\
	constraints(col(l, t, val, cons)) = cons  \\ \\
l \in Label, t = DbType, val = Strings, cons = Constraints
\end{align*}

\paragraph{Foreign key}
\begin{align*}
	\mathbf{ForeignKey} = <label, tableRef, Constraint*> \\ \\
	\mathbf{fk} : Label \times Label \times Constraints \rightarrow ForeignKey \\ \\
	\mathbf{name} : ForeignKey \rightarrow Label \\
	name(fk(l, tRef, cons)) = lab  \\ \\
	\mathbf{reference} : ForeignKey \rightarrow Label  \\
	reference(fk(l, tRef, cons)) = tRef  \\ \\
	\mathbf{constaints} : ForeignKey \rightarrow Constraint*  \\
	constaints(fk(l, tRef, cons)) = cons  \\ \\
l, tRef \in Label, cons \in Constraints
\end{align*}

\paragraph{Primary key}
\begin{align*}
	\mathbf{PrimaryKey} =  < label > 	\\
	\mathbf{pk} : Label \rightarrow PrimaryKey
\end{align*}

\paragraph{DbType} represents primitive types in the database. There are usually defined types such as Varchar, Integer, Boolean etc. There are only two types defined in the model $DBSTRING$ as a default type of values in columns and $DBINT$ as a default type for keys.
\begin{align*}
	\mathbf{DbType} = DBSTRING \; | \; DBINT 
\end{align*}

\paragraph{Constraint} there are two types of constraints defined in the model. Both constraints are column constraints - first constraint defines unique records in a column, second constraint defines non-empty columns.
\begin{align*}
	\mathbf{Constraint} = NOTNULL \; | \; UNIQUE 
\end{align*}

@@

\subsubsection{Model of Stored Data}
A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of reference to a concrete table and a set of value pairs, which represents concrete values of concrete columns.
\begin{align*}
	\mathbf{Row} = < refTable, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, pairs) = l \\ \\
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \in Database
\end{align*}
There functions $contains$ verify if the table contains data in context of concrete database.

\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Label \\ 
	value([l,val]) = val \\ \\
	\mathbf{pairByRef} : Label \times Row \rightarrow Pair \\
	pairByRef(l,i) = \begin{cases}
			\inference{\exists p \in pairs(i) : reference(p) = l}{p} \\ \\
			\varnothing
		\end{cases}\\
	l \in Label, val \in Value
\end{align*}
The $refCol$ is a name of a column, a foreign key or its primary key and $value$ represents concrete value stored in the database. Each row in a database has to refer to a table in database schema and has a structure corresponding to the one defined by a table, otherwise the database is inconsistent. The consistency of database is then:
\begin{align*}
	\mathbf{consistent} : Database \rightarrow Boolean \\
	consistent(d) = \begin{cases}
 		\inference{\forall r \in data(d), \exists t \in tables(d) : reference(r) = name(t) \\ \wedge \forall p \in pairs(r) :\\ \exists u \in columns(t) \cup foreignKeys(t) : \\ reference(p) = name(u)}{true}
 	\\ \\
 	false
 \end{cases}
\end{align*}

@@

\begin{align*}
	\mathbf{addColumn} : Label \times Column* \times Database \rightarrow Database \\ 	
	addColum(l, c, d) = \begin{cases}
		\inference{|c| > 1}{addColumn(l, head(c), d) \bullet addColumn(l, tail(c), d)} \\ \\
		\inference{|c| = 1 \wedge \exists t \in tables(d) \wedge name(t) = l \\ \wedge \forall f \in columns(t) : name(f) \neq name(c)}{columns(t) \cup c}\\ \\
		Inconsistent \;d 
	 \end{cases}\\ 
	 l \in Label, c = Column*, d = Database
\end{align*}

\begin{align*}
	\mathbf{copyColumn} : Label \times Label \times Label \times Database \rightarrow Database \\
	copyColumn(l_s, l_t, l_c, d) = \begin{cases}
 		\inference{\exists t \in tables(d) : name(t) = l_s \\ \wedge \exists c \in columns(t) : name(c) = l_c \\ \wedge \exists e \in tables(d) : name(e) = l_t  \\ \wedge \forall q \in properties(e) : name(q) \neq l_c}{columns(e) \cup c }
	\end{cases} \\
	l_s,l_t,l_c \in Label, d = Database
\end{align*}

\begin{align*}
	\mathbf{addTable} : Table* \times Database \rightarrow Database \\ 
	addTable(t, d) = \begin{cases}
		\inference{|t| > 1}{addTable(head(t),d) \bullet addTable(tail(t), d)}
\\ \\	
		\inference{|t| = 1 \wedge \forall e \in tables(d) \wedge name(t) \neq name(e)}{tables(d) \cup c}
\\ \\
		Inconsistent \;d 
	 \end{cases}\\ 
	t = Table*,  d = Database
\end{align*}

\begin{align*}
	\mathbf{insertRow} : Row* \times Database \rightarrow Database \\
	insertRow(r, d) = \begin{cases}
			\inference{|r|>1}{insertRow(head(r), d) \bullet insertRow(tail(r), d)} \\ \\
		 	\inference{|r|=1 \wedge \exists t \in tables(d) : name(t) = reference(r) }{data(d) \cup r} \\ \\
			 Inconsistent \; d
		 \end{cases}	\\
		 r = Row*, d = Database
\end{align*}

\begin{align*}
	\mathbf{appendValues} : Label \times Pair* \times Database \rightarrow Daatabase \\
	appendValues(l, p, d) = \begin{cases}
 		\inference{|p| > 1}{appendValues(l, head(p), d)\bullet appendValues(l, tail(p), d)}  \\ \\
 		\inference{|p| = 1 \wedge \forall e \in pairs(r_t) : reference(e) \neq reference(p) \\ \wedge \exists t \in tables(d) : name(t) = refernces(r_t), \\ \wedge \exists c \in columns(t) : reference(p) = name(c)}{pairs(r) \cup p} \\ \\
	 	 Inconsistent \; d
 		\end{cases} \\
l \in Label, p = Pair*, d = Databases
\end{align*}

\begin{align*}
	\mathbf{copyValue} : Label \times Label \times Label \times (Instancece \times Instance) \times Database \times Database \\
	copyValue(l_s, l_t, l_{sc}, l_{tc} f_m, d) = \begin{cases}
 		\inference{\exists t_s \in tables(d) : name(t_s) = l_s, \\ t_t \in tables(d) : name(t_t) = l_t, \\ c \in columns(t_s) : name(c) = l_{sc}, \\ e \in columns(t_t) : name(e) = l_{tc} }{\begin{gathered}
			\forall p \in f_m(t_s, t_t)  : \\ appendValues(name(second(p)), \\ [l_{tc},value(pairByRef(l_{sc}, first(p))),d)
		\end{gathered}}
 	\end{cases}
\end{align*}
